\textit{k}-mer-based GWAS enhances the discovery of causal variants and candidate genes in soybean

### 2023-04-04 additional_files/manuscript.tex
Format references for submission to The Plant Genome
Last changes before re-reading manuscript prior to submission; shorten abstract and adjust citations

%%% Load packages
\usepackage{amsthm,amsmath}
\RequirePackage[authoryear]{natbib}% uncomment this for author-year bibliography
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc} %unicode support

% Packages added by M-A Lemay
\usepackage{booktabs} % prettier tables
\usepackage{threeparttable} % easy addition of footnotes below tables
\usepackage[T1]{fontenc}
\usepackage{pdflscape}
\usepackage{graphicx}
\usepackage{ccaption}
\usepackage[margin=1in]{geometry} % leaves more space for text body

@@

%%% Begin ...
\begin{document}

@@

@@

Genome-wide association studies (GWAS) are powerful statistical methods that
detect associations between genotype and phenotype at genome scale. Despite
their power, GWAS frequently fail to pinpoint the causal variant or the gene
controlling a given trait in crop species. Assessing genetic
variants beyond single-nucleotide polymorphisms (SNPs) could alleviate this
problem, for example by including structural variants (SVs).
In this study, we tested the potential of SV- and \emph{k}-mer-based GWAS in
soybean by applying these methods to 13 traits.  We also performed
conventional GWAS analysis based on SNPs and small indels.  We assessed the performance of each GWAS approach based on
results at loci for which the causal genes or variants were known from
previous genetic studies. We found that \emph{k}-mer-based GWAS was the
most versatile approach and the best at pinpointing causal variants or
candidate genes based on the most significantly associated
\emph{k}-mers.  Moreover, \textit{k}-mer-based analyses identified
promising candidate genes for loci related to pod color, pubescence
form, and resistance to \textit{Phytophthora sojae}. In
our dataset, SV-based GWAS did not add value compared to
\textit{k}-mer-based GWAS and may not be worth the time and
computational resources required to genotype SVs at population scale.
Despite promising results, significant challenges remain regarding the
downstream analysis of \emph{k}-mer-based GWAS.  Notably, better
methods are needed to associate significant \emph{k}-mers with sequence
variation. Together, our results suggest that coupling \emph{k}-mer-
and SNP/indel-based GWAS is a powerful approach for discovering
candidate genes in crop species.

@@

\section*{Background}
Genome-wide association studies (GWAS) are analytical approaches that detect
statistical associations between phenotypic observations for a trait of
interest and the genotypes of variants distributed throughout the
genome \cite{tam2019}. This approach has been used for over 15 years
\cite{visscher2012} and has enabled significant advances in our understanding
of the genetic architecture of traits involved in human health
\cite[e.g.][]{wellcome2007gwas} and crop \cite{tibbs2021} and animal
\cite{ma2019} production, among others.

Despite their power and convenience for finding significantly associated loci,
GWAS analyses frequently fail to pinpoint the gene(s) associated with a trait and
the causal variant(s) involved \cite{tam2019}. One possible reason for this
is that many GWAS analyses do not include the causal variant itself in their variant dataset
\cite{tibbs2021}. This could be either because the analysis
is restricted to a particular type of variant, such as single-nucleotide
polymorphisms (SNPs), or because the set of markers/variants genotyped in the
association panel is limited by the technology being used
\cite[e.g.][]{bandillo2015, sonah2015}. Despite this limitation, variants
located near the gene controlling the trait often do appear statistically
associated with the phenotype through linkage disequilibrium (LD) with the
causal variant \cite{korte2013}.  A typical workflow then involves the
identification of candidate genes within the haplotype blocks defined by the
statistically associated markers and follow-up functional analyses to confirm
that a given gene controls the trait \cite[e.g.][]{wang2018, liu2020ps}.

The inclusion of structural variants (SVs) as genotype data in GWAS may partly overcome
this issue by providing a set of variants with a potentially high functional impact.
SVs include any variant involving a difference of at least 50 nucleotides, such as
deletions, insertions, or inversions \cite{ho2020}. Such variants are known to
have large phenotypic impacts by disrupting coding or regulatory sequences
\cite{marroni2014}, and have been identified as causal variants in traits such
as resistance to soybean cyst nematode \cite{cook2012}, aluminum tolerance in
wheat \cite{maron2013}, and branching in maize \cite{studer2011}.  SVs have
already been used in GWAS with promising results
\cite[e.g.][]{zhang2015, akakpo2020, dominguez2020, liu2020-pangenome}, but
their use in GWAS is the exception rather than the norm.

Despite their phenotypic impact, SVs are still largely understudied compared to
SNPs because discovering and genotyping them with accuracy is difficult.
Comprehensive detection of SVs requires a combination of several methods that
often need to be tailored to the study species and data available
\cite{alkan2011, ho2020}. Even then, assessments of the performance of SV
discovery and genotyping generally reveal subpar sensitivity and precision
\cite[e.g.][]{cameron2019, chaisson2019, kosugi2019}. Moreover, despite recent
improvements in computational approaches for genotyping SVs
\cite[e.g.][]{siren2021, ebler2022}, such approaches remain relatively
inaccurate compared to SNP genotyping workflows.

Alternative approaches that allow one to assess a large spectrum
of variants in GWAS without genotyping SVs at population scale could be useful.
Voichek and Weigel \cite{voichek2020} developed a \emph{k}-mer-based GWAS approach that may
represent such an alternative method. This approach relies on the presence or
absence of \emph{k}-mers observed in sequence reads in place of variant genotypes for use in
GWAS. Once a presence/absence table of \emph{k}-mers has been generated for the
population under study, this table can be used in association analyses
similarly to SNP or SV genotype calls. One advantage of using \emph{k}-mers is
that they can act as a molecular signature for any type of variant as long as
these variants result in presence/absence of \emph{k}-mers of a given length.
Another advantage is that \emph{k}-mers are not tied to a specific genomic
location and can thus be used to query associations between genotype and
phenotype at genomic locations that are not included in a reference genome.  A
few studies have applied this \emph{k}-mer-based GWAS approach
\cite[e.g.][]{tripodi2021, colque2021} and other similar approaches have been
developed \cite{rahman2018, he2021}, but similarly to the use of SVs in GWAS,
such approaches have not been widely adopted yet.

In this study, we assessed the potential of SV- and \emph{k}-mer-based GWAS in
a major crop, soybean (\emph{Glycine max}). We also conducted conventional GWAS
based on SNPs and indels. In order to test these methods, we analyzed ten
qualitative traits \cite{bandillo2017} and two quantitative traits
\cite{bandillo2015} that have been previously studied using a SNP array
developed for soybean \cite{song2013}. In addition, we analyzed a quantitative
trait (horizontal resistance to \emph{Phytophthora sojae}) that has been
previously studied using whole-genome sequencing (WGS) data
\cite{deronne2022}.  Since the underlying genes and causal variants are
already known for several of these traits, they provide a good test case for
the performance of GWAS methods.  The objectives of our study were to:

@@

\section*{Methods}

@@

We selected 389 inbred \emph{G. max} accessions based on the availability of
Illumina WGS data and phenotypic data for resistance to the oomycete \emph{P.
sojae} \citep{deronne2022}.  We identified 741 SRA runs corresponding to
Illumina paired-end sequencing data for those 389 accessions \citep[data from][]{zhou2015, valliyodan2016, fang2017, bayer2021}
and extracted paired-end reads using the fastq-dump command v. 2.9.6
\cite{sratoolkit} with the
option \texttt{-\-split-3}.
Reads that were not paired in the raw sequencing data were not used
for downstream analyses.  The runs downloaded from the SRA and associated
metadata are listed in Supplemental file 2.

We first filtered the sequencing data for quality and the presence of sequencing
adapters using bbduk from BBtools v. 38.25
\cite{bbtools}. Reads
were then mapped to the reference assembly version 4 of \emph{G. max}
cultivar Williams82 \citep{valliyodan2019} concatenated with chloroplast
and mitochondrion sequences obtained from SoyBase \citep{grant2010}. We used
bwa-mem v. 0.7.17-r1188 \citep{li2009-bwa} with default parameters to map the
reads in paired-end mode. Reads that were left unpaired following adapter- and
quality-trimming were aligned separately in single-end mode with bwa-mem and
merged with mapped paired-end reads using samtools \texttt{merge}
v. 1.12 \citep{li2009-samtools}.  We then added read groups identifying
individual SRA runs to the BAM files and merged the reads belonging to the same
accession using bamaddrg
\cite{bamaddrg}. We used the
resulting BAM files for all downstream analyses requiring aligned reads.
Metadata on mapping depth of aligned data can be found in Supplemental file 3.

@@

\begin{itemize}
	\item We used Illumina WGS data from the 389 accessions mentioned above
		to call SVs in the population following a pipeline outlined
		by Lemay et al. \cite{lemay2022}. In brief, we combined information from
		the SV calling programs AsmVar \citep{liu2015}, Manta \citep{chen2016},
		smoove \citep{pedersen2019} and SvABA \citep{wala2018} to generate
		a set of candidate SVs from the mapped Illumina data.
	\item We used SVs discovered by Lemay et al. \cite{lemay2022} from Oxford Nanopore
		sequencing data of 17 Canadian soybean cultivars.
	\item We called SVs from 26 high-quality genome assemblies published by
		Liu et al. \cite{liu2020-pangenome}, as well as that of the cultivar Zhonghuang 13 (ZH13),
		\textit{G. soja} accession W05, and \textit{G. max} cultivar Lee.
		We used methods similar to those of Liu et al. \cite{liu2020-pangenome} to call
		SVs from the comparison of these assemblies to that of Williams82.
\end{itemize}

@@

We used the approach outlined by Voichek and Weigel \cite{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\citep{kokot2017}. \emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used \emph{k}-mers with a MAF $\geq$ 0.02 for both the
computation of kinship and GWAS.

@@

We analyzed ten qualitative traits previously studied by Bandillo et al. \cite{bandillo2017}: flower
color, pubescence color, seed coat color, stem termination type, hilum color,
pod color, pubescence form, pubescence density, seed coat luster, and maturity
group.  We obtained phenotypic data for these traits from the GRIN database
(\url{https://npgsweb.ars-grin.gov/gringlobal/search}) by querying the database
using the PI identifiers of the 389 accessions used.  Four out of the 389
accessions did not match the GRIN database and were therefore not used for the
analysis of qualitative traits. The phenotypes of qualitative traits were recoded
to numerical values for GWAS analysis following methods similar to those of
Bandillo et al. \cite{bandillo2017}.  For some traits, we conducted more than one GWAS analysis
using targeted subsets of the observed phenotypes in order to focus on specific
loci. The phenotypes and numerical values used for all analyses as well as the
number of observations in the dataset that we used for GWAS are listed in
Supplemental Tables S1 to S10.

We analyzed three quantitative traits in addition to the ten qualitative traits
mentioned above. Seed oil and protein content were retrieved from the GRIN
database. In cases where more than one value was listed for a given accession,
we computed the average of those values and used it for GWAS analyses. As a third quantitative trait, we
used the horizontal resistance to the oomycete \textit{P. sojae} (corrected dry
weight, CDW) as recently assessed in a hydroponic assay \citep{deronne2022}.
Our analysis included 340 accessions whose phenotypic data
was already published by de Ronne et al. \cite{deronne2022} as well as 49 additional accessions for
which sequencing data was not yet available when their study was conducted.
Because of ambiguity regarding the identity of one of the accessions (HN019), this accession
was dropped from all GWAS analyses. Similarly to seed oil and
protein content, the CDW values for some accessions for which more than one
observation was available were averaged.

@@

We ran GAPIT using a mixed linear model (MLM) with 9 principal components and the VanRaden
algorithm for computing the kinship matrix. We used an MLM model for
consistency with the statistical model used by the \textit{k}-mer-based
approach.  We used a randomization approach to determine the 5\% family-wise
error-rate threshold for each GWAS analysis as described by Voichek and Weigel \cite{voichek2020}.
To compute this threshold, we permuted the phenotypic observations 100 times
and computed a GWAS on these permuted phenotypes to obtain a distribution of
top \emph{p}-values under the null hypothesis. The fifth most significant of
the top 100 \emph{p}-values was used as the significance threshold for
inference, i.e. markers with \emph{p}-values lower than this threshold were
considered significant.

We conducted GWAS based on \emph{k}-mers using the method developed by
Voichek and Weigel \cite{voichek2020}. In brief, the
analysis was conducted in two steps. The first step used an approximate model
to identify 1 million potentially significant \emph{k}-mers. These were then used in
the second step as input to an exact model implemented in GEMMA \citep{zhou2012}, resulting
in a \emph{p}-value for each \emph{k}-mer. We
used the kinship computed from the \emph{k}-mers to correct for relatedness
between accessions. The program automatically outputs a list of \emph{k}-mers
that passed a 5\% family-wise error-rate threshold determined by a
randomization approach, which we considered significant for the purposes of
downstream analyses.

@@

Zabala and Vodkin \cite{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Supplemental Figures
S2 and S3). The most
significant SV and \textit{k}-mers corresponded to the known causal variant
(Figure \ref{flower-color-main-figure}a-b and Supplemental Figure S4c-d).
In particular, the assembly of the reads containing the most significant \textit{k}-mers
and the alignment of the resulting haplotypes showed that these \textit{k}-mers
directly tagged the causal variant (Figure \ref{flower-color-main-figure}c-e).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, the causal variant could not be captured by this approach. Still, SNPs
and indels located within the gene sequence were among the most significantly
associated ones (Supplemental Figure S4b).

@@

Zabala and Vodkin \cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as a putative
causal variant at this locus. For all three GWAS approaches, we found a signal
overlapping this gene (Supplemental Figure S6). In
particular, the regions defined by the 1\% most associated variants
identified by the SNP- and \textit{k}-mer-based approaches included
Glyma.06g202300 (Supplemental Figure S7). While the most
significant \textit{k}-mers did occur within the sequence of this gene, they
did not correspond to the documented causal variant (Supplemental Figure
S8d).  Still, the fourth and fifth most strongly
associated \textit{k}-mers corresponded to this causal variant (Supplemental Figures
S8d and S9). We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide (Supplemental Figure S9c), which could indicate
either that other causal variants are resulting in a non-functional F3’H in
this population, or that the causal variant lies elsewhere. As for SNP-based GWAS,
the most significant variant within the sequence of Glyma.06g202300 was the
causal variant, however it was not the most significantly associated variant
overall (Supplemental Figures S7b and
S8b). SV-based GWAS did not detect the causal variant
at this locus, as expected from its size.

@@

We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all three
approaches (Supplemental Figure S15).
Wang et al. \cite{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant. Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
SNP-based analysis as corresponding to that causal variant (Figure
\ref{seed-coat-color-main-figure} and Supplemental Figures
S16 and
S17). Unsurprisingly, SV-based GWAS
did not identify the causal variant given its type.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the SNP- and
\textit{k}-mer-based analyses performed best since the region defined by their
most highly associated markers (top 1\%) included the causal gene (Supplemental Figures
S18 and S19).
We did identify one of the SNPs suggested as putative causal variants for this
locus by Tian et al. \cite{tian2010} and Liu et al. \cite{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Supplemental Figures
S20 and S21).
Our results are consistent with previous findings in suggesting that there is
more than one causal variant at the \textit{Dt1} locus \citep{liu2010,
tian2010}. Indeed, a few cultivars bearing the allele associated with
\emph{dt1} showed indeterminate or semi-determinate phenotypes and some
cultivars showed a determinate phenotype without bearing the allele associated
with \emph{dt1} (Supplemental Figure S21c). The existence of
several causal variants might explain why our analyses did not identify a
causal variant more clearly.

@@

The difference in size between the CNV reported by Liu et al. \cite{liu2020ps} and our
results may be due to the presence of some unresolved sequence (a stretch of
4,635 N nucleotides) over that interval in the reference sequence that we used.
Because our SV filtering process removed any alternate allele containing N nucleotides,
we reasoned that the CNV may have been found by our pipeline but
filtered out later on. Indeed, we found a 42.8-kb duplication corresponding
to the one described above among the genotype calls made by smoove. An association
analysis of that single variant using GAPIT found a \textit{p}-value of 1.9e-23,
which would have made it the most significant SV. In this particular case, the
failure of the SV-based GWAS analysis to find identify the causal variant was due
to our filtering parameters and not to limitations of the SV genotype calls per se.

@@

\section*{Availability of data and materials}

>>>>

%%% Load packages
\usepackage{amsthm,amsmath}
\usepackage[hidelinks]{hyperref}
\usepackage[utf8]{inputenc} %unicode support

% Packages added by M-A Lemay
\usepackage{booktabs} % prettier tables
\usepackage{threeparttable} % easy addition of footnotes below tables
\usepackage[T1]{fontenc}
\usepackage{pdflscape}
\usepackage{graphicx}
\usepackage{ccaption}
\usepackage[margin=1in]{geometry} % leaves more space for text body
\usepackage[natbibapa]{apacite}

@@

\begin{document}

@@

Received \_\_\_\_\_\_\_\_\_\_\_

@@

Genome-wide association studies (GWAS) are powerful statistical methods that
detect associations between genotype and phenotype at genome scale. Despite
their power, GWAS frequently fail to pinpoint the causal variant or the gene
controlling a given trait in crop species. Assessing genetic variants other than
single-nucleotide polymorphisms (SNPs) could alleviate this problem.  In this
study, we tested the potential of SV- and \emph{k}-mer-based GWAS in soybean by
applying these methods as well as conventional SNP/indel-based GWAS to 13
traits. We assessed the performance of each GWAS approach based on loci for
which the causal genes or variants were known from previous genetic studies. We
found that \emph{k}-mer-based GWAS was the most versatile approach and the best
at pinpointing causal variants or candidate genes.  Moreover,
\textit{k}-mer-based analyses identified promising candidate genes for loci
related to pod color, pubescence form, and resistance to \textit{Phytophthora
sojae}. In our dataset, SV-based GWAS did not add value compared to
\textit{k}-mer-based GWAS and may not be worth the time and computational
resources invested.  Despite promising results, significant challenges remain
regarding the downstream analysis of \emph{k}-mer-based GWAS.  Notably, better
methods are needed to associate significant \emph{k}-mers with sequence
variation. Our results suggest that coupling \emph{k}-mer- and SNP/indel-based
GWAS is a powerful approach for discovering candidate genes in crop species.

@@

\section*{Introduction}
Genome-wide association studies (GWAS) are analytical approaches that detect
statistical associations between phenotypic observations for a trait of
interest and the genotypes of variants distributed throughout the
genome \citep{tam2019}. This approach has been used for over 15 years
\citep{visscher2012} and has enabled significant advances in our understanding
of the genetic architecture of traits involved in human health
\citep[e.g.][]{wellcome2007gwas} and crop \citep{tibbs2021} and animal
\citep{ma2019} production, among others.

Despite their power and convenience for finding significantly associated loci,
GWAS analyses frequently fail to pinpoint the gene(s) associated with a trait and
the causal variant(s) involved \citep{tam2019}. One possible reason for this
is that many GWAS analyses do not include the causal variant itself in their variant dataset
\citep{tibbs2021}. This could be either because the analysis
is restricted to a particular type of variant, such as single-nucleotide
polymorphisms (SNPs), or because the set of markers/variants genotyped in the
association panel is limited by the technology being used
\citep[e.g.][]{bandillo2015, sonah2015}. Despite this limitation, variants
located near the gene controlling the trait often do appear statistically
associated with the phenotype through linkage disequilibrium (LD) with the
causal variant \citep{korte2013}.  A typical workflow then involves the
identification of candidate genes within the haplotype blocks defined by the
statistically associated markers and follow-up functional analyses to confirm
that a given gene controls the trait \citep[e.g.][]{wang2018, liu2020ps}.

The inclusion of structural variants (SVs) as genotype data in GWAS may partly overcome
this issue by providing a set of variants with a potentially high functional impact.
SVs include any variant involving a difference of at least 50 nucleotides, such as
deletions, insertions, or inversions \citep{ho2020}. Such variants are known to
have large phenotypic impacts by disrupting coding or regulatory sequences
\citep{marroni2014}, and have been identified as causal variants in traits such
as resistance to soybean cyst nematode \citep{cook2012}, aluminum tolerance in
wheat \citep{maron2013}, and branching in maize \citep{studer2011}.  SVs have
already been used in GWAS with promising results
\citep[e.g.][]{zhang2015, akakpo2020, dominguez2020, liu2020-pangenome}, but
their use in GWAS is the exception rather than the norm.

Despite their phenotypic impact, SVs are still largely understudied compared to
SNPs because discovering and genotyping them with accuracy is difficult.
Comprehensive detection of SVs requires a combination of several methods that
often need to be tailored to the study species and data available
\citep{alkan2011, ho2020}. Even then, assessments of the performance of SV
discovery and genotyping generally reveal subpar sensitivity and precision
\citep[e.g.][]{cameron2019, chaisson2019, kosugi2019}. Moreover, despite recent
improvements in computational approaches for genotyping SVs
\citep[e.g.][]{siren2021, ebler2022}, such approaches remain relatively
inaccurate compared to SNP genotyping workflows.

Alternative approaches that allow one to assess a large spectrum
of variants in GWAS without genotyping SVs at population scale could be useful.
\cite{voichek2020} developed a \emph{k}-mer-based GWAS approach that may
represent such an alternative method. This approach relies on the presence or
absence of \emph{k}-mers observed in sequence reads in place of variant genotypes for use in
GWAS. Once a presence/absence table of \emph{k}-mers has been generated for the
population under study, this table can be used in association analyses
similarly to SNP or SV genotype calls. One advantage of using \emph{k}-mers is
that they can act as a molecular signature for any type of variant as long as
these variants result in presence/absence of \emph{k}-mers of a given length.
Another advantage is that \emph{k}-mers are not tied to a specific genomic
location and can thus be used to query associations between genotype and
phenotype at genomic locations that are not included in a reference genome.  A
few studies have applied this \emph{k}-mer-based GWAS approach
\citep[e.g.][]{tripodi2021, colque2021} and other similar approaches have been
developed \citep{rahman2018, he2021}, but similarly to the use of SVs in GWAS,
such approaches have not been widely adopted yet.

In this study, we assessed the potential of SV- and \emph{k}-mer-based GWAS in
a major crop, soybean (\emph{Glycine max}). We also conducted conventional GWAS
based on SNPs and indels. In order to test these methods, we analyzed ten
qualitative traits \citep{bandillo2017} and two quantitative traits
\citep{bandillo2015} that have been previously studied using a SNP array
developed for soybean \citep{song2013}. In addition, we analyzed a quantitative
trait (horizontal resistance to \emph{Phytophthora sojae}) that has been
previously studied using whole-genome sequencing (WGS) data
\citep{deronne2022}.  Since the underlying genes and causal variants are
already known for several of these traits, they provide a good test case for
the performance of GWAS methods.  The objectives of our study were to:

@@

\section*{Materials and methods}

@@

We selected 389 inbred \emph{G. max} accessions based on the availability of
Illumina WGS data and phenotypic data for resistance to the oomycete \emph{P.
sojae} \citep{deronne2022}.  We identified 741 SRA runs corresponding to
Illumina paired-end sequencing data for those 389 accessions \citep[data from][]{zhou2015, valliyodan2016, fang2017, bayer2021}
and extracted paired-end reads using the fastq-dump command v. 2.9.6
(\citeauthor{sratoolkit}) with the
option \texttt{-\-split-3}.
Reads that were not paired in the raw sequencing data were not used
for downstream analyses.  The runs downloaded from the SRA and associated
metadata are listed in Supplemental file 2.

We first filtered the sequencing data for quality and the presence of sequencing
adapters using bbduk from BBtools v. 38.25
(\citeauthor{bbtools}). Reads
were then mapped to the reference assembly version 4 of \emph{G. max}
cultivar Williams82 \citep{valliyodan2019} concatenated with chloroplast
and mitochondrion sequences obtained from SoyBase \citep{grant2010}. We used
bwa-mem v. 0.7.17-r1188 \citep{li2009-bwa} with default parameters to map the
reads in paired-end mode. Reads that were left unpaired following adapter- and
quality-trimming were aligned separately in single-end mode with bwa-mem and
merged with mapped paired-end reads using samtools \texttt{merge}
v. 1.12 \citep{li2009-samtools}.  We then added read groups identifying
individual SRA runs to the BAM files and merged the reads belonging to the same
accession using bamaddrg
(\citeauthor{bamaddrg}). We used the
resulting BAM files for all downstream analyses requiring aligned reads.
Metadata on mapping depth of aligned data can be found in Supplemental file 3.

@@

\begin{itemize}
	\item We used Illumina WGS data from the 389 accessions mentioned above
		to call SVs in the population following a pipeline outlined
		by \cite{lemay2022}. In brief, we combined information from
		the SV calling programs AsmVar \citep{liu2015}, Manta \citep{chen2016},
		smoove \citep{pedersen2019} and SvABA \citep{wala2018} to generate
		a set of candidate SVs from the mapped Illumina data.
	\item We used SVs discovered by \cite{lemay2022} from Oxford Nanopore
		sequencing data of 17 Canadian soybean cultivars.
	\item We called SVs from 26 high-quality genome assemblies published by
		\cite{liu2020-pangenome}, as well as that of the cultivar Zhonghuang 13 (ZH13),
		\textit{G. soja} accession W05, and \textit{G. max} cultivar Lee.
		We used methods similar to those of \cite{liu2020-pangenome} to call
		SVs from the comparison of these assemblies to that of Williams82.
\end{itemize}

@@

We used the approach outlined by \cite{voichek2020} for \emph{k}-mer-based GWAS
and their recommendations on the
associated GitHub page (v. 0.2-beta, \url{https://github.com/voichek/kmersGWAS}) to generate
a presence/absence \emph{k}-mer table for use in GWAS.  The first step of
the approach involved counting all \emph{k}-mers of length 31 (31-mers) present
in the trimmed (FASTQ) reads for each accession using KMC3 v. 3.2.1
\citep{kokot2017}. \emph{k}-mers were counted twice, the first time using their
canonized form (the first of either the observed \textit{k}-mer or its reverse
complement in lexicographical order) and the second time using the observed
\emph{k}-mers themselves (non-canonized form). The two sets of \textit{k}-mers
were then combined and filtered by
keeping only those seen in at least 5 accessions, and observed in both canonized
and non-canonized form in at least 20\% of the accessions in which they are found.
We then generated a table indicating the presence or absence of each
\emph{k}-mer in all accessions. This table was used to compute a kinship
table for use in GWAS and was also used directly as input genotypes for the
GWAS analyses. We used \emph{k}-mers with a MAF $\geq$ 0.02 for both the
computation of kinship and GWAS.

@@

We analyzed ten qualitative traits previously studied by \cite{bandillo2017}: flower
color, pubescence color, seed coat color, stem termination type, hilum color,
pod color, pubescence form, pubescence density, seed coat luster, and maturity
group.  We obtained phenotypic data for these traits from the GRIN database
(\url{https://npgsweb.ars-grin.gov/gringlobal/search}) by querying the database
using the PI identifiers of the 389 accessions used.  Four out of the 389
accessions did not match the GRIN database and were therefore not used for the
analysis of qualitative traits. The phenotypes of qualitative traits were recoded
to numerical values for GWAS analysis following methods similar to those of
\cite{bandillo2017}.  For some traits, we conducted more than one GWAS analysis
using targeted subsets of the observed phenotypes in order to focus on specific
loci. The phenotypes and numerical values used for all analyses as well as the
number of observations in the dataset that we used for GWAS are listed in
Supplemental Tables S1 to S10.

We analyzed three quantitative traits in addition to the ten qualitative traits
mentioned above. Seed oil and protein content were retrieved from the GRIN
database. In cases where more than one value was listed for a given accession,
we computed the average of those values and used it for GWAS analyses. As a third quantitative trait, we
used the horizontal resistance to the oomycete \textit{P. sojae} (corrected dry
weight, CDW) as recently assessed in a hydroponic assay \citep{deronne2022}.
Our analysis included 340 accessions whose phenotypic data
was already published by \cite{deronne2022} as well as 49 additional accessions for
which sequencing data was not yet available when their study was conducted.
Because of ambiguity regarding the identity of one of the accessions (HN019), this accession
was dropped from all GWAS analyses. Similarly to seed oil and
protein content, the CDW values for some accessions for which more than one
observation was available were averaged.

@@

We ran GAPIT using a mixed linear model (MLM) with 9 principal components and the VanRaden
algorithm for computing the kinship matrix. We used an MLM model for
consistency with the statistical model used by the \textit{k}-mer-based
approach.  We used a randomization approach to determine the 5\% family-wise
error-rate threshold for each GWAS analysis as described by \cite{voichek2020}.
To compute this threshold, we permuted the phenotypic observations 100 times
and computed a GWAS on these permuted phenotypes to obtain a distribution of
top \emph{p}-values under the null hypothesis. The fifth most significant of
the top 100 \emph{p}-values was used as the significance threshold for
inference, i.e. markers with \emph{p}-values lower than this threshold were
considered significant.

We conducted GWAS based on \emph{k}-mers using the method developed by
\cite{voichek2020}. In brief, the
analysis was conducted in two steps. The first step used an approximate model
to identify 1 million potentially significant \emph{k}-mers. These were then used in
the second step as input to an exact model implemented in GEMMA \citep{zhou2012}, resulting
in a \emph{p}-value for each \emph{k}-mer. We
used the kinship computed from the \emph{k}-mers to correct for relatedness
between accessions. The program automatically outputs a list of \emph{k}-mers
that passed a 5\% family-wise error-rate threshold determined by a
randomization approach, which we considered significant for the purposes of
downstream analyses.

@@

\cite{zabala2007} identified the flavonoid 3’5’-hydroxylase (F3’5’H) gene
Glyma.13g\-072100 as associated with the \textit{W1} locus for flower color. In
Williams82 and other accessions with white flowers, a 65-bp insertion into the
third exon of the gene results in a premature stop codon that renders the
F3’5’H enzyme non-functional and prevents pigmentation of the flower. All
GWAS approaches detected a signal overlapping the gene at this locus (Supplemental Figures
S2 and S3). The most
significant SV and \textit{k}-mers corresponded to the known causal variant
(Figure \ref{flower-color-main-figure}a-b and Supplemental Figure S4c-d).
In particular, the assembly of the reads containing the most significant \textit{k}-mers
and the alignment of the resulting haplotypes showed that these \textit{k}-mers
directly tagged the causal variant (Figure \ref{flower-color-main-figure}c-e).
As the catalogue of variants analyzed by Platypus contains only SNPs and small
indels, the causal variant could not be captured by this approach. Still, SNPs
and indels located within the gene sequence were among the most significantly
associated ones (Supplemental Figure S4b).

@@

\cite{zabala2003} identified Glyma.06g202300 as the gene associated with the
\textit{T} locus for pubescence color and found a 1-bp deletion as a putative
causal variant at this locus. For all three GWAS approaches, we found a signal
overlapping this gene (Supplemental Figure S6). In
particular, the regions defined by the 1\% most associated variants
identified by the SNP- and \textit{k}-mer-based approaches included
Glyma.06g202300 (Supplemental Figure S7). While the most
significant \textit{k}-mers did occur within the sequence of this gene, they
did not correspond to the documented causal variant (Supplemental Figure
S8d).  Still, the fourth and fifth most strongly
associated \textit{k}-mers corresponded to this causal variant (Supplemental Figures
S8d and S9). We
observed that not all accessions with gray pubescence harbored the deleted
nucleotide (Supplemental Figure S9c), which could indicate
either that other causal variants are resulting in a non-functional F3’H in
this population, or that the causal variant lies elsewhere. As for SNP-based GWAS,
the most significant variant within the sequence of Glyma.06g202300 was the
causal variant, however it was not the most significantly associated variant
overall (Supplemental Figures S7b and
S8b). SV-based GWAS did not detect the causal variant
at this locus, as expected from its size.

@@

We conducted a second GWAS analysis contrasting accessions with yellow and
green seed coats in order to target the \textit{G} locus. Accordingly, this
analysis found strong signals overlapping the \emph{G} locus using all three
approaches (Supplemental Figure S15).
\cite{wang2018} identified Glyma.01g198500 as the gene associated with
this locus and found an A>G SNP affecting transcript splicing as the causal
variant. Consistently with their results, we identified the two
most significant \emph{k}-mers and the second most significant marker from the
SNP-based analysis as corresponding to that causal variant (Figure
\ref{seed-coat-color-main-figure} and Supplemental Figures
S16 and
S17). Unsurprisingly, SV-based GWAS
did not identify the causal variant given its type.

@@

While all GWAS approaches found signals near the Glyma.19g194300 gene
associated with the \textit{Dt1} locus for stem termination type, the SNP- and
\textit{k}-mer-based analyses performed best since the region defined by their
most highly associated markers (top 1\%) included the causal gene (Supplemental Figures
S18 and S19).
We did identify one of the SNPs suggested as putative causal variants for this
locus by \cite{tian2010} and \cite{liu2010}, however this SNP was only the
135\textsuperscript{th} most associated marker found by Platypus and the
12\textsuperscript{th} most associated \emph{k}-mer (Supplemental Figures
S20 and S21).
Our results are consistent with previous findings in suggesting that there is
more than one causal variant at the \textit{Dt1} locus \citep{liu2010,
tian2010}. Indeed, a few cultivars bearing the allele associated with
\emph{dt1} showed indeterminate or semi-determinate phenotypes and some
cultivars showed a determinate phenotype without bearing the allele associated
with \emph{dt1} (Supplemental Figure S21c). The existence of
several causal variants might explain why our analyses did not identify a
causal variant more clearly.

@@

The difference in size between the CNV reported by \cite{liu2020ps} and our
results may be due to the presence of some unresolved sequence (a stretch of
4,635 N nucleotides) over that interval in the reference sequence that we used.
Because our SV filtering process removed any alternate allele containing N nucleotides,
we reasoned that the CNV may have been found by our pipeline but
filtered out later on. Indeed, we found a 42.8-kb duplication corresponding
to the one described above among the genotype calls made by smoove. An association
analysis of that single variant using GAPIT found a \textit{p}-value of 1.9e-23,
which would have made it the most significant SV. In this particular case, the
failure of the SV-based GWAS analysis to find identify the causal variant was due
to our filtering parameters and not to limitations of the SV genotype calls per se.

@@

\section*{Data availability}
